Is malaria control profitable? Return on investment of privately-managed residential fumigations at a large sugarcane processing facility in Southern Mozambique

### 2018-04-20 inst/rmd/paper.tex
major overhaul

Joe Brew\footremember{isglobal}{Barcelona Institute for Global Health: c/ Rosselló, 132, 5è 2a. 08036, Barcelona, Spain, Spain}\footremember{cism}{Centro de Investigação em Saúde de Manhiça: Vila da Manhiça, Bairro Cambeve, Rua 12, Distrito da Manhiça, CP 1929, Maputo, Mozambique, Mozambique}\footremember{vu}{VU University Amsterdam: De Boelelaan 1105, 1081 HV Amsterdam, Netherlands, Netherlands} \footnote{Corresponding Author}
Kizito Gondo\footremember{ma}{Maragra Açucar SA, Subsidiary of Illovo Sugar Ltd: CP 2789, Maputo, Mozambique, Mozambique}
Elton Dorkin\footrecall{ma}
Menno Pradhan\footrecall{vu}\footremember{uva}{University of Amsterdam: REC E, Roetersstraat 11, Amsterdam, Netherlands, Netherlands}
Laia Cirera\footrecall{isglobal}\footrecall{cism}
Ranjeeta Thomas\footremember{icl}{Imperial College London: South Kensington Campus, London SW7 2AZ, U.K., UK}
Elisa Sicuri\footrecall{isglobal}\footrecall{cism}\footrecall{icl}

@@

From the persepctive of the private sector, however, investing in
malaria control is not so clear-cut. Public health interventions
targetting malaria - and their corresponding cost-effectiveness
evaluations - most often focus on impacts pertaining to public welfare,
such as an increase in life years adjusted for disability or quality
(Goodman et al., 1999) (Shretta et al., 2016) (Lee et al., 2017)
(Hanson, 2004). Though population-level health is certainly of
importance to businesses, and improvements in health incidentally
improve the economy at all levels (Brundtland, 1999) (Bloom and Canning,
2008) (Vecchi et al., 2013), these improvements may be too disperse or
long-term to incentivize private sector involvement in health campaigns.

@@

A significant sector of the economy in Mozambique is dominated by
large-scale foreign direct investment projects (Robbins and Perkins,
2012), and the role of the private sector in health generally, and
malaria specifically, is unequivocally important. Large agriculture and
extractive industry firms take up wide swaths of land and employ
hundreds of thousands (German et al., 2013). The Mozambican state has
encouraged large-scale entreprise with the aim of general economic
development (Buur et al., 2012). And where large firms exist, they often
take on social roles such as housing and health care (Winkler, 2013). At
times, this role is necessary from a purely practical standpoint; in
other cases, it is employed under the auspices of ``corporate social
responsibility'' (Azemar and Desbordes, 2009). Regardless of the
language used, it is clear that private industry plays an important role
in public health in Mozambique (Robbins and Perkins, 2012)
(Castel-Branco, 2014).

Several cases exist of foreign firms engaging in large-scale malaria
control campaigns (Mouzin and al., 2011) (Han, 2015) (Bennett et al.,
2017) (Kaula et al., 2017). But these studies generally consider
population health as the outcome measure of interest, rather than worker
absenteeism or productivity. Similarly, they often neglect to
differentiate between those clinical costs which are absorbed by the
local health system versus those which are absorbed by the firm itself.
In the literature, making the ``investment case'' for malaria control or
elimination generally implies that the investor is the public sector,
and takes into account those costs and benefits which are applicable
from a public welfare point of view (Shretta et al., 2017); though
appropriate in most cases (the government or institutions interested in
public welfare primarily being the primary malaria control agents in
most locations), the findings of these studies are rarely applicable to
the private sector, and even less so at granular levels (such as an
individual firm). In the case of a private firm not interested in
``corporate social responsibility'', it is not clear whether investing
in malaria control would be profitable or not. This lack of clarity not
only discourages investment, but also makes it difficult for governments
to pinpoint the correct of amount of subsidy (if applicable) to
encourage private sector scale-up in malaria control.

To address the question of the profitability of malaria control
activities from the standpoint of a private firm, we analyze data during
a 4 year period from a private sugar facility in Southern Mozambique. We
assess the effect of indoor residual spraying (IRS) on the absenteeism
and health of workers, and demonstrate that the firm's engagement in
malaria control not only improved worker health, but also generated a
positive return on investment from a pure accounting perspective.

The structure of the paper is as follows. After a brief review of the
existing literature on the subject of private sector investment in
malaria control and its effects on worker productivity, we provide an
overview of the sugar company under study, and the epidemiology of
malaria in the nearby area, as well as in Mozambique as a whole. We then
give an overview of the data collected, and outline the theoretical and
methdological assumptions and tools that underly our analysis. In the
results, we show the effectiveness of the company's malaria control
program in reducing absences, and translate this reduction into cost
savings. Our discussion covers potential implications from this study in
terms of policy and investment, as well as the paper's limitations.

@@

Regardless of the academic literature on the risk of workers becoming
infected with malaria, the companies running large agricultural
operations in malaria-endemic regions generally consider the disease to
be an important enough threat to workers' health to merit some form of
investment in malaria control (Pluess et al., 2009). Insecticide
fumigation, bednet distribution, test-and-treat clinics, and occasional
educational campaigns are common forms of corporate social
responsibility to combat malaria. A World Bank study on Nigerian
sugarcane cutters found that the simple availability of testing and
treatment increased productivity by 10\% in the weeks following the
provision of services, the conclusion being that both the treatment and
the test result were effective in increasing productivity, the latter
simply increasing the information available which could influence
personal labor allocation decisions (Dillon et al., 2014).

Malaria may disincentivize investment in business in malaria zones. In a
vicious cycle, businesses already in malaria zones may not be
incentivized to invest in malaria control, perhaps because this is seen
as the domain of the public sector (Curtis et al., 2003), or perhaps due
to not knowing the extent to which the benefits in health of that
control will accrue to the business itself.

@@

\begin{figure}[H]

@@

The mill of the Maragra Açucar SA (a subsidary of the Illovo sugar
company, henceforth referred to as ``Maragra'') was nationalized in the
1970s (like all other Mozambican mills), went through a period of low
production, and then fell completely out of use by 1984. It re-opened in
private hands in 1992, and was renovated by a group of international
investors in 1998. Today, Maragra accounts for roughly one quarter of
Mozambique's overall sugar production (second only to the nearby
Xinavane mill run by the Hulett Sugar Tongaat company) (Sutton, 2014).
With a favorably close location to the port of Maputo, ample land,
approximately 5,000 employees (of which three quarters are seasonal),
and a mill with the capacity to process not only all the sugarcane grown
on Maragra's land, but also the cane of the many nearby smallholders
(O'Laughlin, 2016), Maragra has so far been able to weather the 2016
Mozambican crisis and concurrent collapse in global sugar prices.

@@

\begin{figure}[H]

{\centering \includegraphics{figures/unnamed-chunk-14-1}

@@

\caption{i. Fumigation activities carried out by Maragra Malaria Control during study period, ii. Distribution of average time between sprayings of households}\label{fig:unnamed-chunk-14}
\end{figure}

@@

Maragra regularly employs IRS at on-site worker households in order to
reduce those workers' (and their families') risk of malaria infection.
Workers living off-site (our control group) also may have received IRS
at some point during the study period (from government programs). Even
though we do not have reliable person-level data on IRS carried out by
the government, off-site workers are a suitable control in the sense
that they represent ``business as usual'' (ie, what would happen if the
company carried out no IRS and relied solely on public interventions).
Using company HR and clinical records, we were able to identify absences
and episodes of clinical malaria among all workers, as well as identify
the time since the most recent IRS episode before the onsent of absence
or illness.

@@

We sought to understand the effect of IRS on individual workers'
likelihood of absence from work as well as their likelihood of clinical
malaria. To estimate this effect, we estimated separate models for
absence and illness. We employed linear mixed effects models (Minalu et
al., 2011) (Liu et al., 2010) (with individual worker fixed effects, and
random effects for other covariates) in lieu of interrupted time series
(Lopez Bernal et al., 2016) so as to explicitly adjust for confounders
(Bell and Jones, 2014). We divide into 4 different models for different
worker types, so as to account for the potential time-varying effects of
worker type on risk of malaria. Our econometric model is relatively
straightforward:

\[
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{it}) + (\beta_2{IRS}*\beta_3{IRS_it}) + ... + \epsilon_{it}
\]

\(\hat{Y}\) is the rate of absence. \(\beta_{1}\) represents the
clinical malaria incidence at that time in the entire district of
Manhiça. Our demographic confounders (represented by \(...\)) are sex,
age, and worker department (field, factory, or administrative). Our
intervention whether the residence of the worker in question was treated
in the last year, and, if so, the time since treatment (represented
above as the interaction term, where where \(_t\) represents time
elapsed since commencement of the most recent IRS campaign). We define
the malaria season as any time during which the clinical incidence of
malaria in the district of Manhiça was at or greater than the median
clinical incidence of malaria for the entire study period. These weeks
are flagged as red in Figure 1, Panel A. By using clinical incidence of
the area of residence of the workers (as opposed to more typical proxies
for malaria risk, such as only rainy vs.~non rainy season), our
seasonality estimate is a closer approximation of true malaria risk,
incorporating lagged effects such as the incubation period of the
parasite, as well as any inherent non-linear effects of weather. In
addition, we adjust for daily precipitation

@@

We collected absenteeism and demographic data data on 3362 workers from
2012 through 2016. Workers were approximately 60\% male, and more than
80\% fieldworkers (predominantly cane-cutters). Most were in their 20s
and 30s and employed on temporary contracts. Table 1 shows overall
worker details.

\begin{table}[ht]
\centering
\begin{tabular}{rlll}
  \hline
Variable &   & Workers & (\%) \\ 
  \hline
Department & Administrative & 157 & 4.67 \\ 
   & Factory & 459 & 13.65 \\ 
   & Field & 2746 & 81.68 \\ 
  Sex & F & 1305 & 38.82 \\ 
   & M & 2057 & 61.18 \\ 
  Status & Permanent & 500 & 14.87 \\ 
   & Temporary & 2862 & 85.13 \\ 
  Age & 20 & 985 & 29.30 \\ 
   & 30 & 1244 & 37.00 \\ 
   & 40 & 665 & 19.78 \\ 
   & 50 & 347 & 10.32 \\ 
   & 60 & 117 & 3.48 \\ 
   & 70 &   4 & 0.12 \\ 
   \hline
\end{tabular}
\caption{Overall worker characteristics} 
\end{table}

The overall absenteeism rate during the study period was 5.56\% among
all workers. The number of malaria cases registered at the company
clinic during that time was 1873, for an approximate annualized clinical
incidence of 110 cases per 1,000.

@@

\begin{figure}[H]

{\centering \includegraphics{figures/unnamed-chunk-17-1}

\caption{Age distribution by sex and worker status}\label{fig:unnamed-chunk-17}
\end{figure}

\begin{figure}[H]

@@

\caption{i. Monthly total precipitation in the Maniça district; ii. Average monthly temperature (bars) during the study period, as well as monthly maximum (red) and minimum (blue) temperatures}\label{fig:unnamed-chunk-18}
\end{figure}

@@

In Southern Mozambique, malaria peaks during the summer months (December
through March) most years (Figure 4, panel A), and worker absenteeism at
Maragra rates track malaria incidence closely, following the same
seasonal patterns (figure 4, panel B). Both all-cause absenteeism and
sick absenteeism have declined in recent years at Maragra (figure 4,
panel C), with the latter declining at a faster rate than the former.
The fact that the rate of confirmed cases at the company clinic is
largely non-seasonal (figure 4, panels E and F) suggests that a
significant portion of workers either seek care for malaria elsewhere
(for example, government health posts, of which several are nearby and
in some many cases closer to workers' residence than the company clinic)
or do not seek care during malaria infection. Accordingly, we focus our
analysis on all-cause absenteeism rather than sickness absenteeism or
malaria diagnostics, with the assumption that much of illness is
captured by absenteeism but not by the clinical data.

\begin{figure}
\centering
\includegraphics{figures/unnamed-chunk-19-1.pdf}
\caption{Clinical malaria (district of Manhiça), all-cause absenteeism
among Maragra workers, sick absenteeism among Maragra workers, positive
cases at company clinic, and test positivity rate at company clinic}
\end{figure}

\textbf{Fumigations}: During the period from January 1st, 2012 through
December 31st, 2016, the Maragra Malaria Control Unit carried out 11,567
episodes of fumigation of residential ``agregados'' (household
combinations), for a total of 13,937 building-fumigation combinations.
The total number of unique agregados sprayed during this period was
4,045. Among the 3,362 workers for whom we have reliable absenteeism and
residential data, 692 had their homes fumigated at least once (the
majority of workers live off of the facility).

@@

\textbf{Effect of IRS on clinical malaria}: xxx.
\todo{Not done yet. In fact, not even sure if I should do this section at all, since it's basically just a distraction.}

@@

\begin{figure}
\centering
\includegraphics{figures/unnamed-chunk-22-1.pdf}
\caption{i. Absenteeism before and after IRS administration for all
workers who ever received IRS; ii. The same, but segregated by malaria
and non-malaria seasons}
\end{figure}

In order to assess the effect of IRS on absenteeism, we created 4 worker
fixed effects models for the 4 principal worker types (permanent field
worker, temporary field worker, permanent non-field worker, temporary
non-field worker). The reason for segregating models rather than
incorporating worker type as a fixed effect was that it seemed plausible
that the effect of the treatment (IRS) on the outcome would be different
(and vary differently over time) depending on these worker types. For
instance, it is reasonable to assume that IRS' effect would be
differential for a temporary worker (who likely lives in the fumigated
house less than 100\% of the year) than for a permanent worker.

@@

The model results (table 3) suggest a significant decrease in
absenteeism following the administration of IRS. The effect is
ambivalent among temporary workers, whereas among permanent workers the
administration of IRS decreases absenteeism betwen 2.5 and 8.4
percentage points in the 2-9 month following administration. These
effects are consistent with IRS decreasing absence through the medium of
malaria, since (a) the effect is not as strong in the first month
(potentially during the parasite incubation phase), the effect is
stronger on field workers (who are more socioeconomically the
demographic at greater risk of malaria than their administrative
counterparts), and the effect is ambivalent among temporary workers
(since the protective effect of IRS is only a function of the amount of
time that the worker spends in the fumigated house).

@@

\textbf{Translating absences to costs}

@@

\begin{itemize}
\item
  Overview of findings
\item
  How this contributes to the literature
\item
  Implications for policy and for businesses
\end{itemize}

@@

@@

@@

@@

>>>>

Joe Brew\footremember{isglobal}{Barcelona Institute for Global Health: c/ Rosselló, 132, 5è 2a. 08036, Barcelona, Spain, Spain}\footremember{cism}{Centro de Investigação em Saúde de Manhiça: Vila da Manhiça, Bairro Cambeve, Rua 12, Distrito da Manhiça, CP 1929, Maputo, Mozambique, Mozambique}\footremember{vu}{VU University Amsterdam: De Boelelaan 1105, 1081 HV Amsterdam, Netherlands, Netherlands} \footnote{Corresponding Author}
Kizito Gondo\footremember{ma}{Maragra Açucar SA, Subsidiary of Illovo Sugar Ltd: CP 2789, Maputo, Mozambique, Mozambique}
Elton Dorkin\footrecall{ma}
Menno Pradhan\footrecall{vu}\footremember{uva}{University of Amsterdam: REC E, Roetersstraat 11, Amsterdam, Netherlands, Netherlands}
Laia Cirera\footrecall{isglobal}\footrecall{cism}
Elisa Sicuri\footrecall{isglobal}\footrecall{cism}\footremember{icl}{Imperial College London: South Kensington Campus, London SW7 2AZ, U.K., UK}

@@

From the persepctive of the private sector, however, investing in
malaria control is not so clear-cut, since the benefits are often
disperse, long-term, and difficult to quantify. Public health
interventions targetting malaria - and their corresponding
cost-effectiveness evaluations - most often focus on impacts pertaining
to public welfare, such as an increase in life years adjusted for
disability or quality (Goodman et al., 1999) (Shretta et al., 2016) (Lee
et al., 2017) (Hanson, 2004). Though population-level health is
certainly of importance to businesses, and improvements in health
incidentally improve the economy at all levels (Brundtland, 1999) (Bloom
and Canning, 2008) (Vecchi et al., 2013), these improvements may be too
disperse or long-term to incentivize private sector involvement in
health campaigns. In other words, the returns for malaria control are
less for the private sector because (i) they capture only part of the
benefits and (ii) they do not benefit from the externalities.

Just as the benefits of malaria control to the private sector are more
limited than those to the public sector, considerations regarding costs
for a firm are also distinct than those for a government. Though many
firms in endemic regions engage in malaria control programs, this should
not be considered, per se, evidence of its cost-effectiveness (since the
extent to which corporate social responsibility plays a role is
unknown).

@@

A significant sector of the economy in Mozambique is dominated by
large-scale foreign direct investment projects (Robbins and Perkins,
2012), and the role of the private sector in health generally, and
malaria specifically, is unequivocally important. Large agriculture and
extractive industry firms take up wide swaths of land and employ
hundreds of thousands (German et al., 2013). The Mozambican state has
encouraged large-scale entreprise with the aim of general economic
development (Buur et al., 2012). And where large firms exist, they often
take on social roles such as housing and health care (Winkler, 2013). At
times, this role is necessary from a purely practical standpoint; in
other cases, it is employed under the auspices of ``corporate social
responsibility'' (Azemar and Desbordes, 2009) (Curtis et al., 2003).
Regardless of the language used, it is clear that private industry plays
an important role in public health in Mozambique (Robbins and Perkins,
2012) (Castel-Branco, 2014).

To address the question of the profitability of malaria control
activities from the standpoint of a private firm, we analyze data during
a 4 year period from a private sugar facility in Southern Mozambique. We
use absenteeism as our outcome variable, assuming that it is directly
correlated with the productivity losses associated with malaria
infection. We assess the effect of indoor residual spraying (IRS) on
absenteeism, and demonstrate that the firm's engagement in malaria
control not only improved worker health, but also generated a positive
return on investment from a pure accounting perspective.

The structure of the paper is as follows. We provide an overview of the
sugar company under study, and the epidemiology of malaria in the nearby
area, as well as in Mozambique as a whole. We then give an overview of
the data collected, and outline the theoretical and methdological
assumptions and tools that underly our analysis. We assess the effect of
time since IRS on worker absenteeism, controlling for malaria
seasonality, and segregating models for four different worker types. In
the results, we show that IRS spraying is associated with a significant
reduction in worker absenteeism among permanent and fieldworkers, and
has little to no effect on temporary and indoor workers. We find that,
in addition to reducing absence, the IRS program has a cost savings
effect. Our discussion covers potential implications from this study in
terms of policy and investment, as well as the paper's limitations.

\subsection{Background literature}\label{background-literature}

In general, large firms operating in malaria endemic regions consider
malaria to be an important enough issue to merit at least some
investment (Pluess et al., 2009). Several studies examine the effect of
foreign firms engaging in large-scale malaria control campaigns (Han,
2015) (Bennett et al., 2017) (Kaula et al., 2017). AngloGold-Ashanti, in
partnership with local and national government in Ghana, invested in a
well-rounded malaria control program in 2005, and saw worker absenteeism
fall by 50\% in 13 months (CCM, 2016). Lafarge's simulatenous investment
in a comprehensive malaria control program in Benin was associated with
an average 41\% reduction in absenteeism among workers over the course
of 4 years (Egedeye et al., 2011). Zambia Sugar Plc, Zambia's largest
sugar processing facility, saw annual malaria cases at its company
clinic fall from nearly 3,000 in 2001 to less than 500 by 2005,
following investment in a malaria control program. Marathon Oil's
investment of 15 million US in vector control, education, net
distribution and malaria treatment on Bioko Island in Equatorial Guinea
lead to an estimated 95\% reduction in the number of parasite-infected
mosquitoes and 50\% reduction in malaria incidence among young children
(Asquino, 2016) (Overgaard et al., 2012). A PATH study in Zambia found a
return on investment of 28\% among three companies investing in
employer-based malaria control (Mouzin and al., 2011).
\todo{Still need to include stuff from http://gbchealth.org/wp-content/uploads/2014/03/IRS2011WorkshopReport.pdf}

Though certainly suggestive of high returns on malaria control
investments, these studies generally consider population health as the
outcome measure of interest, rather than worker absenteeism or
productivity. Similarly, they often neglect to differentiate between
those clinical costs which are absorbed by the local health system
versus those which are absorbed by the firm itself. When absenteeism
itself is considered, effects of malaria control have generally been
found to be high, but causation is difficult to establish, given that
the previous studies rely on aggregate data.

Two studies utilize worker-level data to estimate the effect of malaria
control on productivity. A World Bank analysis of Nigerian sugarcane
cutters found that the simple availability of testing and treatment
increased productivity by 10\% in the weeks following the provision of
services, the conclusion being that both the treatment and the test
result were effective in increasing productivity, the latter simply
increasing the information available which could influence personal
labor allocation decisions (Dillon et al., 2014). A randomized
controlled trial (RCT) in Zambia showed an even greater effect from
investments in preventing malaria: farmers given bed nets saw fewer days
lost to illness (both directly and due to caretaking responsibilities
for ill family members), translating to an increase of approximately
15\% in crop yield (Fink and Masiye, 2015). Though compelling, the
Nigerian program only dealt with medical services (diagnostics and
medication), rather than preventive interventions, and the Zambian RCT
examined individual farmers, as opposed to a large firm.

In the literature, making the ``investment case'' for malaria control or
elimination generally implies that the investor is the public sector,
and takes into account those costs and benefits which are applicable
from a public welfare point of view (Shretta et al., 2017). For example,
an economic analysis by the Corporate Alliance on Malaria in Africa on
the Bioko Island Malaria Control Program found a 4:1 cost-benefit ratio,
but the perspective in this case was considered to be the ``community''
(Egedeye et al., 2011). Though appropriate in most cases to consider
benefits accrued to the community (the government or institutions
interested in public welfare primarily being the primary malaria control
agents in most locations), the findings of these studies are rarely
applicable to the private sector, and even less so at granular levels
(such as an individual firm). In the case of a private firm not
interested in ``corporate social responsibility'', it is not clear
whether investing in malaria control would be profitable or not. This
lack of clarity not only discourages investment, but also makes it
difficult for governments to pinpoint the correct of amount of subsidy
(if applicable) to encourage private sector scale-up in malaria control.

@@

@@

\begin{figure}[h]

@@

The mill of the Maragra Açucar SA (a subsidary of the Illovo sugar
company, henceforth referred to as ``Maragra'') was nationalized in the
1970s (like all other Mozambican mills), went through a period of low
production, and then fell completely out of use by 1984. It re-opened in
private hands in 1992, and was renovated by a group of international
investors in 1998. Today, Maragra accounts for roughly one quarter of
Mozambique's overall sugar production (second only to the nearby
Xinavane mill run by the Hulett Sugar Tongaat company) (Sutton, 2014).
With a favorably close location to the port of Maputo, ample land
(approximately 90 squared kilometers of plantation, and 5 squared
kilometers of factory grounds), approximately 5,000 employees (of which
three quarters are seasonal), and a mill with the capacity to process
not only all the sugarcane grown on Maragra's land, but also the cane of
the many nearby smallholders (O'Laughlin, 2016), Maragra has so far been
able to weather the 2016 Mozambican crisis and concurrent collapse in
global sugar prices.

@@

\begin{figure}[h]

{\centering \includegraphics{figures/unnamed-chunk-15-1}

@@

\caption{i. Fumigation activities carried out by Maragra Malaria Control during study period, ii. Distribution of average time between sprayings of households}\label{fig:unnamed-chunk-15}
\end{figure}

@@

Maragra regularly employs IRS at on-site worker households in order to
reduce those workers' (and their families') risk of malaria infection.
IRS works by killing the malaria vector (mosquito), thereby preventing
infection of the household occupants. When administered correctly, IRS
is a low-risk intervention to its recipients, and is assumed not to
affect absenteeism in the short-term (to the extent that it may cause
negative side-effects, these are generally long-term). It is preventive
only, and does not cure current malaria infection, nor does it affect
the parasite load of mosquitoes. Workers living off-site (our control
group) also may have received IRS at some point during the study period
(from government programs). Even though we do not have reliable
person-level data on IRS carried out by the government, off-site workers
are a suitable control in the sense that they represent ``business as
usual'' (ie, what would happen if the company carried out no IRS and
relied solely on public interventions). Using company HR and clinical
records, we were able to identify absences and episodes of clinical
malaria among all workers, as well as identify the time since the most
recent IRS episode before the onsent of absence or illness.

@@

We sought to understand the effect of IRS on individual workers'
likelihood of absence from work as well as their likelihood of clinical
malaria. To estimate this effect, we estimated separate models for
absence and illness. We employed linear mixed effects models (Minalu et
al., 2011) (Liu et al., 2010) (with individual worker fixed effects, and
random effects for other covariates) in lieu of interrupted time series
(Lopez Bernal et al., 2016) so as to explicitly adjust for confounders
(Bell and Jones, 2014). We divide into 4 different models for different
worker types, so as to account for the potential time-varying effects of
worker type on risk of malaria. Our model specification is as follows

\[
\hat{Y_{it}} = \hat{\beta}_{0} +  \hat{\beta}_{1}\text{Season}_{t} * (\hat{\beta}_2{IRS_{it}}*\hat{\beta}_3{IRStime_{it}}) + \alpha_i + \delta_t + \upsilon_{it}
\]

\(\hat{Y}_{it}\) is the rate of absence. \(\beta_{1}\) is the binary
``season'' variable, imputed from overall district clinical incidence.
Our intervention is whether the residence of the worker in question was
treated in the last year, and, if so, the time since treatment,
represented, respectively, by \(\beta_{2}\) and \(\beta_{3}\).
\(\alpha_i\) reprsents the time invariant worker fixed effects, and
\(\delta_i\) represents the fixed effect of the particular malaria
season. \(\upsilon\) is the error term.

We define the malaria season as any time during which the clinical
incidence of malaria in the district of Manhiça was at or greater than
the median clinical incidence of malaria for the entire study period.
These weeks are flagged as red in Figure 1, Panel A. By using clinical
incidence of the area of residence of the workers (as opposed to more
typical proxies for malaria risk, such as only rainy vs.~non rainy
season), our seasonality estimate is a closer approximation of true
malaria risk, incorporating lagged effects such as the incubation period
of the parasite, as well as any inherent non-linear effects of weather.
In addition, we adjust for daily precipitation; though its lagged effect
on malaria incidence is likely captured by the seasonality term, we
include rainfall since it's immediate effect (through its impact on
transportation and working conditions) may also affect absenteeism.

@@

We collected absenteeism and demographic data data on 3362 workers from
2012 through 2016. Workers were approximately 60\% male, and more than
80\% fieldworkers (predominantly cane-cutters). Most were in their 20s
and 30s and employed on temporary contracts. Table 1 shows overall
worker details divided by ``Treatment'' versus ``Control'' status.
``Treatment'' is considered the time beginning at IRS administration and
ending one year later; ``Control'' is considered time observed either
prior to IRS administration (and greater than 1 year after previous IRS
administration) or among workers who never received IRS. Because of the
temporal nature of these categories, many workers belong to both
treatment and control groups, albeit at different times.

\begin{table}[ht]
\centering
\begin{tabular}{rlllll}
  \hline
Variable &   & Treatment workers & Treatment days & Control workers & Control days \\ 
  \hline
Department & Administrative &  35 & 12523 & 157 & 173151 \\ 
   & Factory &  71 & 20533 & 453 & 280416 \\ 
   & Field & 410 & 91302 & 2691 & 1181175 \\ 
  Sex & F & 228 & 47907 & 1274 & 489484 \\ 
   & M & 288 & 76451 & 2027 & 1145258 \\ 
  Status & Permanent & 127 & 47075 & 500 & 655311 \\ 
   & Temporary & 389 & 77283 & 2801 & 979431 \\ 
  Age & 20 & 140 & 29463 & 960 & 347758 \\ 
   & 30 & 176 & 39119 & 1225 & 550783 \\ 
   & 40 & 123 & 30186 & 652 & 366936 \\ 
   & 50 &  58 & 18881 & 344 & 268877 \\ 
   & 60 &  17 & 5893 & 116 & 96555 \\ 
   & 70 &   2 & 816 &   4 & 3833 \\ 
   \hline
\end{tabular}
\caption{Overall worker characteristics} 
\end{table}

@@

\begin{figure}[h]

@@

\caption{Age distribution by sex and worker status}\label{fig:unnamed-chunk-18}
\end{figure}

\begin{figure}[h]

{\centering \includegraphics{figures/unnamed-chunk-19-1}

\caption{i. Monthly total precipitation in the Maniça district; ii. Average monthly temperature (bars) during the study period, as well as monthly maximum (red) and minimum (blue) temperatures}\label{fig:unnamed-chunk-19}
\end{figure}

@@

In Southern Mozambique, malaria peaks during the summer months (December
through March) most years (Figure 5, panel A), and worker absenteeism at
Maragra rates track malaria incidence closely, following the same
seasonal patterns (figure 4, panel B). Both all-cause absenteeism and
sick absenteeism have declined in recent years at Maragra (figure 5,
panel C), with the latter declining at a faster rate than the former.
The fact that the rate of confirmed cases at the company clinic is
largely non-seasonal (figure 5, panels E and F) suggests that a
significant portion of workers either seek care for malaria elsewhere
(for example, government health posts, of which several are nearby and
in some many cases closer to workers' residence than the company clinic)
or do not seek care during malaria infection. Accordingly, we focus our
analysis on all-cause absenteeism rather than sickness absenteeism or
malaria diagnostics, with the assumption that much of illness is
captured by absenteeism but not by the clinical data.

\begin{figure}
\centering
\includegraphics{figures/unnamed-chunk-20-1.pdf}
\caption{Clinical malaria (district of Manhiça), all-cause absenteeism
among Maragra workers, sick absenteeism among Maragra workers, positive
cases at company clinic, and test positivity rate at company clinic}
\end{figure}

\textbf{Fumigations}: During the period from January 1st, 2013 through
December 31st, 2016, the Maragra Malaria Control Unit carried out 11,007
episodes of fumigation of residential ``agregados'' (household
combinations), for a total of 13,260 building-fumigation combinations.
The total number of unique agregados sprayed during this period was
3,998. Among the 3,362 workers for whom we have reliable absenteeism and
residential data, 692 had their homes fumigated at least once (the
majority of workers live off of the facility).

@@

\textbf{Precipitation}: Of the 1454 days observed, 940 had no rainfall
at all (ie, approximately two-thirds). On days for which there was any
rainfall at all, the average amount was approximately 2.99 centimeters.
Rain was most common in December and January (average of 5-5.5 cm daily)
and least common in August and September (average of 0.5 cm daily). Days
with any rainfall saw more absenteeism than days with no rainfall
(Figure 6, panel A) and among days with any rainfall, more precipitation
was associated with greater absenteeism (Figure 6, panel B) (correlation
coefficient of 0.25).

\begin{figure}[h]

{\centering \includegraphics{figures/unnamed-chunk-25-1}

\caption{Rainfall and absenteeism: association of any rainfall with absenteeism rate (left) and association of rainfall amount and absenteeism (right)}\label{fig:unnamed-chunk-25}
\end{figure}

The extent to which rainfall's effect on absenteeism is confounded by
its effect on malaria is discussed in the model results section.

\textbf{Effect of IRS on clinical malaria}: The number of malaria cases
registered at the company clinic during that time was 1873, for an
approximate annualized clinical incidence of 110 cases per 1,000.
\todo{Not done yet. In fact, not even sure if I should do this section at all, since it's basically just a distraction.}

@@

\begin{figure}
\centering
\includegraphics{figures/unnamed-chunk-26-1.pdf}
\caption{i. Absenteeism before and after IRS administration for all
workers who ever received IRS; ii. The same, but segregated by malaria
and non-malaria seasons}
\end{figure}

In order to assess the effect of IRS on absenteeism, we created 4 worker
fixed effects models for the 4 principal worker types (permanent field
worker, temporary field worker, permanent non-field worker, temporary
non-field worker) for both outcoms (all cause absenteeism vs.~sickness
absenteeism). The reason for segregating models rather than
incorporating worker type as a fixed effect was that it seemed plausible
that the effect of the treatment (IRS) on the outcome would be different
(and vary differently over time) depending on these worker types. For
instance, it is reasonable to assume that IRS' effect would be
differential for a temporary worker (who likely lives in the fumigated
house less than 100\% of the year) than for a permanent worker.

@@

The model results for all cause absenteeism (table 4) suggest a
significant decrease in absenteeism following the administration of IRS.
The effect is ambivalent among temporary workers, whereas among
permanent workers the administration of IRS decreases absenteeism betwen
2.5 and 8.4 percentage points in the 2-9 month following administration.
These effects are consistent with IRS decreasing absence through the
medium of malaria, since (a) the effect is not as strong in the first
month (potentially during the parasite incubation phase), the effect is
stronger on field workers (who are more socioeconomically the
demographic at greater risk of malaria than their administrative
counterparts), and the effect is ambivalent among temporary workers
(since the protective effect of IRS is only a function of the amount of
time that the worker spends in the fumigated house).

@@

This section is in progress. Looking at neighborhood effects,
core-periphery effects, etc.

\subsection{Translating absences to
costs}\label{translating-absences-to-costs}

@@

\begin{itemize}
\tightlist
\item
  Overview of findings
\item
  How this contributes to the literature
\item
  Implications for policy and for businesses
\item
  Spillover effects !
\item
  How absenteeism might be \emph{underestimating} true effect, since at
  the margin a sick worker might go to work (ie, not be absent), but be
  less productive
\end{itemize}

@@

\hypertarget{ref-asquino2015}{}
Asquino, M., 2016. Equatorial Guinea Plays a Leading Role in Combating
Malaria.

@@

\hypertarget{ref-anglo}{}
CCM, 2016. AngloGold Ashanti Malaria Control Ltd (AGA Mal).

@@

\hypertarget{ref-lafarge}{}
Egedeye, L., Drozer, S., Leiser, A.-M., 2011. Corporate Action on
Malaria Control: Best Practices and Interventions.

\hypertarget{ref-Fink2015}{}
Fink, G., Masiye, F., 2015. Health and agricultural productivity:
Evidence from zambia. Journal of Health Economics 42, 151--164.
\url{https://doi.org/10.1016/j.jhealeco.2015.04.004}

@@

\hypertarget{ref-Overgaard2012}{}
Overgaard, H.J., Reddy, V.P., Abaga, S., Matias, A., Reddy, M.R.,
Kulkarni, V., Schwabe, C., Segura, L., Kleinschmidt, I., Slotman, M.A.,
2012. Malaria transmission after five years of vector control on bioko
island, equatorial guinea. Parasites \& Vectors 5, 253.
\url{https://doi.org/10.1186/1756-3305-5-253}
